---
title: "Filtering Gene Expression Data"
output: html_document
---
  
## Introduction  
  
You have received a massive dataset containing thousands of genes from an experiment (Influenza A vs. Non-Infected cells).

Your goal is to find the specific genes that changed significantly due to the infection.

To do this, you will need to fill in the missing code (marked as `____`) to filter the data.  
Note: The code blocks below will cause an error if you try to run them before replacing the `____` with your answers!
  
***

### Setting Up Your Workspace

R scripts run in a specific folder on your computer called the **working directory**. Before we begin, we need to know where R is currently looking, and then tell it where it *should* look to find your data.

**How to find your file path:**
* **MAC:** Right-click + Option your file -> select "Copy as Pathname"
* **WINDOWS:** Shift + Right-click your file -> select "Copy as path" (Remember to change `\` to `/` in R!)

```{r setup-directory}
# The getwd() function tells you exactly which folder R is currently looking at.
getwd()

# The setwd() function tells R where your data is saved.
# setwd("C:/Users/YourName/My_file_location") # Windows example
# setwd("/Users/YourName/My_file_location") # Mac example
setwd("/Users/electratsaglioti/Downloads")
```

***

### STEP 0: How to use Code Chunks

Check out this example!

```{r load-data-example, eval=FALSE}
# Replace the ____ with "Hello World"
print(____)

```
  
When you fill it in, it will look like:

```{r load-data-example-completed, eval=FALSE}
print("Hello World")

```

**Rule to remember:** In R, when you are typing "plain text" (like a file name), you must put it in "quotation marks". When you are typing a number or a column name, you don't need to.
  
### STEP 1: Load and Clean the Data
  
First, we need to load the data into R and remove any rows with missing data (NAs). Run the chunk below to get started.

```{r load-data, eval=FALSE}
# Set your working directory (Make sure to change this to your actual folder!)
setwd(____)

# Load the data
# Use `read.table()` to read your text file into a table. 
#`header = TRUE` tells R that the first row contains the column names.
#`sep = "\t"` tells R that the columns are separated by "tabs".
my_data <- read.table("InfluenzaAvsNonInfected.complete.txt", header = TRUE, sep = "\t")

# Remove rows with NAs
# Certain genes had very low expression levels (essentially not expressed), so statistics were not calculated for them, resulting in "NA" (Not Available). The `na.omit()` function scans your data, and if any row has an "NA" in any column, it deletes that row. 
clean_data <- na.omit(my_data)

# You can check both tables from the Environment window to your right.

# Print the names of the columns to the console (will be helpful later..)
colnames(clean_data)

```

***
  
### STEP 2: Filter for Statistical Significance
  
Now we want to find genes that are statistically significant.

**QUESTION 2:** Which column name should you use to check for significance? 

**QUESTION 3:** What is a standard scientific cutoff value for significance (e.g. 0.13)? 


**CODE CHALLENGE:** Replace the `____` sections in the code below with your answers from Questions 2 and 3 to filter the dataset.

```{r filter-significance, eval=FALSE}
# Replace the blanks to keep only statistically significant genes
# The `subset()` function keeps only the rows of clean_data that meet our rule.
sig_genes <- subset(clean_data, ____ < ____)

# Let's count how many we found!
nrow(sig_genes)
```

***
  
### STEP 3: Filter for Biologically Significant UP-Regulated Genes
  
Just because a gene is statistically significant, it doesn't mean that the change is large enough to be biologically significant.

**QUESTION 4:** Look at the columns again. Which columns tell you how much the gene's expression actually changed between the two conditions? 

**QUESTION 5:** Which one is best suited for filtering?

**QUESTION 6:** If a gene is UP-regulated (increased expression), should the numbers in this column be positive or negative? What cutoff value will you choose to represent a "large" increase?

**CODE CHALLENGE:** Fill in the blanks below. You need to include BOTH the rule for your statistical significance AND the rule for your biological significance.

```{r filter-up-regulated, eval=FALSE}
# Filter for UP-regulated genes
# We use the & symbol to combine rules! In our case, we want to combine the statistical rule with the biological significance rule!
up_genes <- subset(clean_data, ____ < ____ & ____ > ____)

# How many UP-regulated genes did you find?
nrow(up_genes)
```

***
  
### STEP 4: Filter for Biologically Significant DOWN-Regulated Genes
  
Now let's find the genes that the virus suppressed!

**CODE CHALLENGE:** Fill in the blanks below to find the down-regulated genes.

```{r filter-down-regulated, eval=FALSE}
# Filter for DOWN-regulated genes
down_genes <- subset(clean_data, ____ < ____ & ____ < ____)

# How many DOWN-regulated genes did you find?
nrow(down_genes)
```

***

### STEP 5: Filter the total statistically and biologically significant genes.
  
Tip: the `abs()`(absolute value) function effectively turns all negative numbers into positive numbers, while positive numbers stay exactly as they are..

**CODE CHALLENGE:** Fill in the blanks below to find the down-regulated genes.

```{r filter-all-degs, eval=FALSE}
# Filter for DOWN-regulated genes
down_genes <- subset(clean_data, ____ < ____ & abs(____) < ____)

# How many DOWN-regulated genes did you find?
nrow(total_genes)
```

### STEP 5: Save Your Discoveries!

You found the genes! Now let's save them to your computer so you can use them for your g:profiler analysis!

**CODE CHALLENGE:** Fill in the `____` with the name(s) of the R object(s) that you want to save into (a) file(s).

```{r save-data, eval=FALSE}
# Saving into a file, for more than one files, copy the code again and change the object name and the file name.
# Remember to always place file names into quotes! (e.g., "my_awesome_results.txt")
write.table(____, file = ____, sep = "\t", row.names = FALSE)


```

***

**Great job! You have successfully filtered an RNA-seq dataset.**